Concept of a Double Tilted Rowland Spectrograph for X-rays

### 2024-02-26 papers/Doubletiltconcept/doubletiltconcept.tex
Doubletilt paper to first round of Co-authors

%% The default is a single spaced, 10 point font, single spaced article.
%% There are 5 other style options available via an optional argument. They
%% can be invoked like this:
%%
%% \documentclass[arguments]{aastex631}
%% 
%% where the layout options are:
%%
%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.  
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%% 		  wider left and right margins. This uses the Daniel
%% 		  Foreman-Mackey and David Hogg design.
%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts 
%%                but now that abstracts are required this is obsolete for
%%                AAS Journals. Authors might need it for other reasons. DO NOT
%%                use \begin{abstract} and \end{abstract} with this style.
%%
%% Note that you can submit to the AAS Journals in any of these 6 styles.
%%
%% There are other optional arguments one can invoke to allow other stylistic
%% actions. The available options are:
%%
%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
%%   tighten      : Makes baselineskip slightly smaller, only works with 
%%                  the twocolumn substyle.
%%   times        : uses times font instead of the default
%%   linenumbers  : turn on lineno package.
%%   trackchanges : required to see the revision mark up and print its output
%%   longauthor   : Do not use the more compressed footnote style (default) for 
%%                  the author/collaboration/affiliations. Instead print all
%%                  affiliation information after each name. Creates a much 
%%                  longer author list but may be desirable for short 
%%                  author papers.
%% twocolappendix : make 2 column appendix.
%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
%%                  for dual anonymous review.
%%
%% these can be used in any combination, e.g.
%%
%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
%%
%% AASTeX v6.* now includes \hyperref support. While we have built in specific
%% defaults into the classfile you can manually override them with the
%% \hypersetup command. For example,
%%
%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
%%
%% will change the color of the internal links to red, the links to the
%% bibliography to green, the file links to cyan, and the external links to
%% magenta. Additional information on \hyperref options can be found here:
%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
%%
%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
%% to improve the accessibility of the compiled pdf file.
%%
%% If you want to create your own macros, you can do so
%% using \newcommand. Your macros should appear before
%% the \begin{document} command.
%%
\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}

@@

@@

\begin{abstract}

This example manuscript is intended to serve as a tutorial and template for
authors to use when writing their own AAS Journal articles. The manuscript
includes a history of \aastex\ and documents the new features in the
previous versions as well as the bug fixes in version 6.31. This
manuscript includes many figure and table examples to illustrate these new
features.  Information on features not explicitly mentioned in the article
can be viewed in the manuscript comments or more extensive online
documentation. Authors are welcome replace the text, tables, figures, and
bibliography with their own and submit the resulting manuscript to the AAS
Journals peer review system.  The first lesson in the tutorial is to remind
authors that the AAS Journals, the Astrophysical Journal (ApJ), the
Astrophysical Journal Letters (ApJL), the Astronomical Journal (AJ), and
the Planetary Science Journal (PSJ) all have a 250 word limit for the 
abstract\footnote{Abstracts for Research Notes of the American Astronomical 
Society (RNAAS) are limited to 150 words}.  If you exceed this length the
Editorial office will ask you to shorten it. This abstract has 182 words.

@@

%% Keywords should appear after the \end{abstract} command. 
%% The AAS Journals now uses Unified Astronomy Thesaurus concepts:
%% https://astrothesaurus.org
%% You will be asked to selected these concepts during the submission process
%% but this old "keyword" functionality is maintained in case authors want
%% to include these concepts in their preprints.
\keywords{Classical Novae (251) --- Ultraviolet astronomy(1736) --- History of astronomy(1868) --- Interdisciplinary astronomy(804)}

@@

\section{Introduction} \label{sec:intro}
X-ray spectroscopy can deliver many important insights into the astrophysics of the hot and energetic universe that are not otherwise observable. In particular, high-resolution spectroscopy (with resolving power $> 500$ or better) can resolve closely-spaced emission lines which offer diagnostics such as density and temperature, reveal the near-edge fine structure in absorption lines, or allow us to detect the precense of faint and narrow absorption lines caused by the warm-hot intergalactic medium in front of bright background quasars to give just a few examples. In the X-ray band, such high resolving powers can be achieved with either microcalorimaters or dispersive elements such as crystals or diffraction gratings. Microcalorimeters offer a fixed absolute energy resolution $\Delta E$, which gives them good resolving power $R=E/\Delta E$ at large energies $E$ and consequently microcalorimeter detectors are planned for the imaging cameras for several of the next X-ray observatories\cite{charm,athena,lynx papers for detectors). However, diffraction gratings have a distinct advantage for softer X-rays below about 2~keV.

The two largest X-ray observatories in history both field diffraction gratings: The Reflection Grating Spectrometer (RGS) on XMM-Newton\cite{RGS} and both the High-Energy Transmission Grating Spectrograph (HETGS) and the Low-Energy Transmission Spectrograph (LETGS) on Chandra\cite{Carnizares2005,LETG}.
**Are each of these called spectrograph or spectrometer?**
Unlike microcalorimeters, diffraction gratings require an additional dispersive element (a transmission or a reflection grating) in the optical path of the instrument and dedicated detectors to capture the dispersed signal. One common layout for a diffraction spectrometer is the Rowlandtorus\cite{Beuermann:78}. In the plan defined bt the optical axis and the dispersion direction, the gratings are located at one side of a circle, the detectors on the other. This circle is commonly called the ``Rowland circle''. In three dimensions, the surface that defines the best grating positions, make up the surface of a torus, thus the name ``Rowland torus'' for this geometry (see Sect.~\ref{sect:onetorus} for a figure).

In this article, we present an optical layout for an instrument that stacks two or four different channels with distinct optical axes in such a way that each channel follows the Rowland torus prescription, yet at the same time all channels are imaged onto a common set of detectors. The general concept could be applied for a large range of focal lengths, grating constants, and other hardware properties. Thus, we discuss advantanges and challenges of this layout without constraint to any specific mission concept. However, we note that Arcus\cite{}, an X-ray spectroscopy mission selected for phase A study in a NASA midex (medium explorer) call in 2018 applied this layout. Specific details of the  Arcus performace, such as predicted effective area and resolving power are shown \citenum{}, but the general concept of a double tilted Rowland spectrograph with its advantages and drawbacks has not been discussed -- a void we attempt to fill with this article.

The layout of this article is as follows: First, we discuss the properties on the optical elements that are most influential for the design of a DTRS, namely mirror properties, gratings, and detectors in Sect.~\ref{sect:elements}. Then, the illustrate the optical layout, modifying the design of a traditional single-channel Rowland torus spectrograph to a DTRS. In section~\ref{sect:discussion}, we discuss advantages and disadvantages of our DTRS concept. We end with a short summart in section \ref{sect:summary}.

\section{Elements of an X-ray grating spectrograph}
\label{sect:elements}

\subsection{Mirror characteristics and sub-aperturing}

\subsection{Dispersive elements}
In prinicle works for any transmission grating, but makes most sense for CAT gratings, so discuss those

trasmission vs diffraction gratings - Casey 2020

Because X-ray can penetrate most material commonly used to manufacture gratings, X-ray instruments are designed to use the gratings either in transmission or, e.g.\ in the case of the RGS, at very small reflection angles close to the angle of total reflection. Thus, transmission and reflection gratings would be located at essentially the same position in the optical path.

\subsection{Detectors}
The dispersed spectrum is spread out far in the dispersion direction. Thus, the ideal detector is much longer in one direction than in the other. Individual detector elements (e.g.\ CCDs) can be tiled, but to avoid chip gaps, it is beneficial to keep the number of elements low. In our DTRS design several spectral traces are positioned in parallel (section~\ref{sect:opticallayout), so the detector has to be position sensitive in 2D to separate individual channels.  Because ir covers a large area than in direct imaging observations, a detector with moderate cost and power requirements per surface area is well-suited for this design. The detector needs to be slightly curved to follow the Rowland torus. If it is made up of smaller individually flat elements, those elements can be arranged tangential to a circle.

The dispersed spectrum contains fewer photons than direct imaging with the same collecting area and spreads this signal out over a large physical region. Thus, the counts rates are lower. Detectors that might by at risk of pile-up in direct observations can still be used for the dispersed signal. On the other hand, the larger regions needed to extract the signal means that background events (internal background and diffuse astrophysical emission) are more important than in direct imaging observations. Similarly, fast timing resolution is typically not needed, because the lower signal means that the detected photons need to be binned up into longer time bins for analysis than a direct imaging observation. Intrinsic energy resolution is required to both suppress the background by filtering out events that are incompatible with wavelengths of the dispersed signal at any particular location and to separate grating orders, where photons of different orders are diffracted to the same position (e.g.\ the sixth order for the O~{\sc vii} forbidden line of the He-like triplet at 2.21~nm and the seventh order of 1.94~mn photons, close to the O~{\sc viii} Lyman$\alpha$ line).

In our DTRS design, the direct images of all channels are positioned on detectors to provide an accurate wavelength calibration, but also because the high-enery part of the spectrum, which is not efficiently dispersed by gratings can provide valuable context for the scientific interpretation of the high-resolution grating spectrum. Thus, the dector type chosen should also be able to handle direct imaging, albeit at a much lower count rate than in a direct imaging instrument without dispersive elements.

**Ask Eric is he knows a paper to cite for comparison of detector types. Lynx?**

\section{Optical Layout}
In this section, we describe the optical layout for a DTRS. We start from the commonly used layout with a single spectrsocopy channel as it is employed in Chandra/LETGS or XMM/RGS. We then show how this design can be modified step-by-step and what considerations are important when chosing the specific parameters for a design. We illustrastrate the concept with ray-trace simulations. Those simulations are performed using a real ray-trace with the marxs code, but the dispersion angle (which is set by the energy of the photons and the grating constant of the grating) is chosen to be larger than feasible for a real instrument. This is done purely to separate the direct light (zero order) and the dispersed light (positive and negative orders) better in the images, so that the concept can be explained more clearly. Static images are embedded in the article, but we strongly encourage the reader explore the interactive 3D models embedded in the electronic version of the publication (might require Acrobat reader).

f, d, sketches

\subsection{One Rowland Torus with central optical axis}
\label{sect:onetorus}
chip gaps chandra, dither

%\begin{figure}
%  \begin{interactive}{js}{file_wth_arll_components.tar/zip}
%  figure call (e.g.\plotone, \includegraphics, etc.)
%  \end{interactive}
%  \caption{Description of the example figure and the interactive portion which includes how to use the interactive functionality.}
%\end{figure}

@@

This torus is tilted Ralfs tild paper
\cite{doi:10.1117/12.856482}

\subsection{Double tilted Rowland Torus}
For R > r signal dips below plane, fish form, two points, symmetry

two opticla axes,
offset chip gaps
spectroscopic vs imganing focus unless I;ve said that above.

\subsection{Channel offsets}

\subsubsection{Channel offsets in the cross-dispersion direction}

\subsubsection{Channel offsets in the dispersion direction}

\section{Discussion}

\subsection{Subaperturing space considerations, and resolving power}
lower mirror requirements, thus chepaer cost for same resolving power, but can use wdge only, stack channels to get square aperture and get back effective area

\subsection{Number of CCDs}
2 camerase, different directions. Need to get zeroth orders anyway (high-energy, wavelength scale), but more CCDs than just one directions (Lynx), make good use of rea-estate

\subsection{Robustness}
relaxes alignment because multiple channels (thus lower cost and technical risk), resilience because independent channels on different CCDS (see XMM loss of RGS chip), check for systemics in calibration because independent channels, chip gaps

wavelength self-calibration (does not need DTRS, but works there, too). is order sorting and several orders ad one energy

\subsection{Other advantages}

Zero order seen, for direct observation (hard X-rays) and calibration

\subsection{Drawbacks}
increased background due ot larger CCD region, less source-free region for background.instrum analysis, coplexity of analysis (numberof spectra!), sprectra coupled (cross-dispersion, read-out streaks for traditional CCDs, wings and energy resolution), does not work for crowded fields, fields with extended emission (galactic center)

>>>>

%% The default is a single spaced, 10 point font, single spaced article.
%% There are 5 other style options available via an optional argument. They
%% can be invoked like this:
%%
%% \documentclass[arguments]{aastex631}
%%
%% where the layout options are:
%%
%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%% 		  wider left and right margins. This uses the Daniel
%% 		  Foreman-Mackey and David Hogg design.
%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts
%%                but now that abstracts are required this is obsolete for
%%                AAS Journals. Authors might need it for other reasons. DO NOT
%%                use \begin{abstract} and \end{abstract} with this style.
%%
%% Note that you can submit to the AAS Journals in any of these 6 styles.
%%
%% There are other optional arguments one can invoke to allow other stylistic
%% actions. The available options are:
%%
%%   astrosymb    : Loads Astrosymb font and define \astrocommands.
%%   tighten      : Makes baselineskip slightly smaller, only works with
%%                  the twocolumn substyle.
%%   times        : uses times font instead of the default
%%   linenumbers  : turn on lineno package.
%%   trackchanges : required to see the revision mark up and print its output
%%   longauthor   : Do not use the more compressed footnote style (default) for
%%                  the author/collaboration/affiliations. Instead print all
%%                  affiliation information after each name. Creates a much
%%                  longer author list but may be desirable for short
%%                  author papers.
%% twocolappendix : make 2 column appendix.
%%   anonymous    : Do not show the authors, affiliations and acknowledgments
%%                  for dual anonymous review.
%%
%% these can be used in any combination, e.g.
%%
%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
%%
%% AASTeX v6.* now includes \hyperref support. While we have built in specific
%% defaults into the classfile you can manually override them with the
%% \hypersetup command. For example,
%%
%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
%%
%% will change the color of the internal links to red, the links to the
%% bibliography to green, the file links to cyan, and the external links to
%% magenta. Additional information on \hyperref options can be found here:
%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
%%
%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
%% to improve the accessibility of the compiled pdf file.
%%
%% If you want to create your own macros, you can do so
%% using \newcommand. Your macros should appear before
%% the \begin{document} command.
%%

@@

\usepackage{amsmath}

@@

\begin{abstract}
High-resolution spectroscopy in soft X-rays ($<2$~keV) requires diffractive elements to resolve any astrophysically relevant diagnostics, such as closely spaced lines, weak absorption lines, or line profiles. For transmission gratings, the Rowland torus geometry describes how gratings and detectors need to be positioned to optimize the spectral resolving power. We describe how an on-axis Rowland geometry can be tilted to accommodate blazed gratings. In this geometry, two channels with separate optical axes can share the same detectors (double tilted Rowland spectrograph, DTRS). Small offsets between the channels can mitigate the effect of chip gaps and reduce the alignment requirements during the construction of the instrument. The DTRS concept is especially useful for sub-apertured mirrors, because it allows an effective use of space in the entrance aperture of a spacecraft.
One mission that applies this concept is the Arcus probe.

@@

%% Keywords should appear after the \end{abstract} command.
%% The AAS Journals now uses Unified Astronomy Thesaurus concepts:
%% https://astrothesaurus.org
%% You will be asked to selected these concepts during the submission process
%% but this old "keyword" functionality is maintained in case authors want
%% to include these concepts in their preprints.
\keywords{Astronomical instrumentation (799) --- Spectrometers (1554) --- X-ray astronomy (1810)}

@@

\section{Introduction} \label{sec:intro}
X-ray spectroscopy can deliver many insights into the astrophysics of the hot and energetic universe that are not otherwise observable. In particular, high-resolution spectroscopy (with resolving power $> 500$ or better) can resolve closely-spaced emission lines which offer diagnostics such as density and temperature, reveal the near-edge fine structure in absorption lines, or allow us to detect the presence of faint and narrow absorption lines caused by the warm-hot intergalactic medium in front of bright background quasars to give just a few examples. In the X-ray band, such high resolving powers can be achieved with either microcalorimaters or dispersive elements such as crystals or diffraction gratings. Microcalorimeters offer a fixed absolute energy resolution $\Delta E$, which gives them good resolving power $R=E/\Delta E$ at large energies $E$ and consequently a microcalorimeter has been launched on Astro-E/Hitomi/SXS \citep{2014SPIE.9144E..2AM} and XRISM/Resolve \citep{2018JATIS...4a1214K} and this technology is also considered for other large missions in development such as Athena \citep{2014SPIE.9144E..2LR} and Lynx \citep{2019JATIS...5b1017B}. \citet{2016Natur.535..117H} used the microcalorimeter data from Astro-E/Hitomi to measure the line-of-sight velocity dispersion of \ion{Fe}{26} in the Perseus cluster and found that the turbulent pressure is only a few percent of the thermodynamic pressure in that region. However, diffraction gratings have a distinct advantage for softer X-rays below about 2~keV.

The two largest X-ray observatories in history both field diffraction gratings: The Reflection Grating Spectrometer (RGS) on XMM-Newton \citep{2001A&A...365L...7D} and both the High-Energy Transmission Grating  \citep[HETG,][]{2005PASP..117.1144C} and the Low-Energy Transmission Grating \citep[LETG,][]{1997SPIE.3113..172P} on Chandra.
Unlike microcalorimeters, diffraction gratings require an additional dispersive element (a transmission or a reflection grating) in the optical path of the instrument and dedicated detectors to capture the dispersed signal. One common layout for a diffraction spectrometer is the Rowlandtorus \citep{Beuermann:78}. In the plane defined by the optical axis and the dispersion direction, the gratings are located at one side of a circle, the detectors on the other. This circle is commonly called the ``Rowland circle''. In three dimensions, the surface that defines the best grating positions has the shape of a torus, thus the name ``Rowland torus'' for this geometry (see Sect.~\ref{sect:onetorus} for a figure).

In this article, we present an optical layout for an instrument that stacks two or four different channels with distinct optical axes in such a way that each channel follows the Rowland torus prescription, yet at the same time all channels are imaged onto a common set of detectors. The general concept could be applied for a large range of focal lengths, grating constants, and other hardware properties. Thus, we discuss advantages and challenges of this layout in general. However, we note that Arcus \citep{2023SPIE12678E..0ES}, an X-ray spectroscopy mission proposed for NASA midex (medium explorer) and probe calls, applies this layout. Specific details of the  Arcus performance, such as predicted effective area and resolving power are shown in \citet{2018SPIE10699E..6FG,2023SPIE12678E..1DG}, but the general concept of a double tilted Rowland spectrograph (DTRS) with its advantages and drawbacks has not been discussed -- a void we attempt to fill with this article.

The structure of this article is as follows: First, we discuss the properties on the optical elements that are most influential for the design of a DTRS, namely mirror properties, gratings, and detectors in Sect.~\ref{sect:elements}. Then, we illustrate the optical layout, modifying the design of a traditional single-channel Rowland torus spectrograph to a DTRS. In section~\ref{sect:discussion}, we discuss advantages and disadvantages of our DTRS concept in general and for the example of Arcus in particular (section~\ref{sect:applications}). We end with a short summary in section \ref{sect:summary}.

\section{Elements of an X-ray grating spectrograph}
\label{sect:elements}
An X-ray spectrograph using the Rowland layout consists of a focussing mirror, diffractive gratings, and a detector.

\subsection{Mirror characteristics and sub-aperturing}
The concept of a DTRS can be implemented with different mirror technologies. The size of the collecting area is to first order proportional to the number of photons that the spectrograph can detect, and the spectral resolving power $\Delta \lambda / \lambda$ increases when the mirror point-spread-function (PSF) decreases. Depending on the type of mirror chosen, the PSF can be dominated by the alignment error between mirror shells, in particular for mirrors that consist of hundreds of elements. In this case, X-rays are offset from the focus point both in the plane of reflection and perpendicular to it.
On the other hand, mirrors can also have a figure error, which is the deviation of the mirror surface from the ideal shape. If the PSF is dominated by the figure error or scattering by particulates on the mirror surface, then scatter in the plane of reflection is typically larger than the out-of-plane scatter. In this case, sub-aperturing of the aperture can improve the spectral resolving power. This is illustrated in Fig.~\ref{fig:subaperture}.

\begin{figure*}
\plotone{subaperture.pdf}
\caption{
    Illustration of the concept of sub-aperturing. All panels show a view from the universe looking down onto the telescope. \emph{left}: Each dot is a photon shown as it passes through a diffraction gratings. Photons are colored according to their polar angle in this panel. \emph{middle and right:} Photons detected in the zeroth and first diffraction order on the detector, respectively. The dispersion direction is from left to right ($y$ coordinate), the distance from the focal point at $(y, z) = (0, 0)$ is shown. Each photon is shown in the same color as in the left panel. Photons passing mirror and grating close to $z=0$ (dark purple) are spread out along the dispersion direction ($y$ coordinate); photons passing mirror and gratings close to $y=0$ are more concentrated along the dispersion direction. Sub-aperturing means that the mirror and gratings do not cover a full circle, but only regions marked in yellow and green in the left panel, so that the resulting photon distributions on the detector are narrower in dispersion direction. Sub-aperturing thus reduces the PSF in the dispersion direction, which increases the spectral resolving power.}
\label{fig:subaperture}
\end{figure*}

\subsection{Dispersive elements}
In a Rowland torus geometry, diffraction gratings are placed on the surface of the Rowland torus. They diffract light according to the grating equation for photons arriving perpendicular to the grating surface:
\begin{equation}
n \lambda = d \sin \theta \label{eqn:diffraction}
\end{equation}
where $\lambda$ is the wavelength of the light, $n$ is the order of the diffraction, $d$ is the grating constant, and $\theta$ is the angle between the incoming light and the normal to the grating. (The equation can be rewritten for rays hitting the grating at arbitrary angles, but for the purpose of this conceptual discussion that is not needed.)
Typically, these transmission gratings are flat and only a few cm$^2$ in size; the spacing between gratings bars $d$ is the same for all gratings and does not vary within a grating. While a flat grating necessarily deviates from the curved surface of a torus, in today's instruments those deviations are small enough that they do not impact the resolving power of the spectrograph significantly. Larger gratings would increase the effective area of the instrument because less space is needed for frames and other mounting structures that hold the gratings in place, but they would also deviate more from the surface of the torus. \citet{2020SPIE11444E..88G} explored bending the gratings in one dimension or varying the distance between grating bars over a grating to compensate for the deviations from the ideal shape in simulations. However, these concepts have not been implemented in a hardware yet.

\subsection{Detectors}
The dispersed spectrum is spread out far in the dispersion direction. Thus, the ideal detector is much longer in one direction than in the other. Individual detector elements (e.g.\ CCDs) can be tiled, but to avoid chip gaps where photons are lost and gaps appear in the extracted spectum, it is beneficial to keep the number of elements low. In our DTRS design several spectral traces are positioned in parallel (section~\ref{sect:opticallayout}), so the detector has to be position sensitive in 2D to separate individual channels.  Because it covers a larger area than in direct imaging observations, a detector with moderate cost and power requirements per surface area is well-suited for this design. The detector should follow the shape of the Rowland circle. This can be achieved by using curved detectors or by tiling smaller flat detector elements where each element is tangential to the Rowland circle.

The gratings spread out the signal over a large physical region on the detector. Thus, the counts rates per pixel are much lower than they would be in an imaging observation. Detectors that might be at risk of pile-up in direct observations can still be used for the dispersed signal. On the other hand, the larger regions needed to extract the signal imply that background events (internal background and diffuse astrophysical emission) are more important than in direct imaging observations. Similarly, fast timing resolution is typically not needed, because the lower signal means that the detected photons need to be binned up into longer time bins for analysis anyway. Intrinsic energy resolution is required to both suppress the background by filtering out events that are incompatible with wavelengths of the dispersed signal at any particular location and to separate grating orders, where photons of different orders are diffracted to the same position (e.g.\ the sixth order for the O~{\sc vii} forbidden line of the He-like triplet at 2.21~nm and the seventh order of 1.94~mn photons, close to the O~{\sc viii} Lyman$\alpha$ line end up at almost the same $\theta$ according to equation~\ref{eqn:diffraction}).

In our DTRS design, the direct images of all channels are positioned on detectors to provide an accurate wavelength calibration, but also because the high-energy part of the spectrum, which is not efficiently dispersed by gratings, can provide valuable context for the scientific interpretation of the high-resolution grating spectrum. Thus, the detector type chosen should also be able to handle direct imaging, albeit at a lower count rate than in a direct imaging instrument without dispersive elements.

\section{Optical Layout}
\label{sect:opticallayout}
In this section, we describe the optical layout for a DTRS. We start from a single central Rowland torus as in Chandra/HETGS and Chandra/LETGS. We then show how this design can be modified step-by-step and what considerations are important when choosing the specific parameters for a design. We illustrate the concept with ray-trace simulations. Those simulations are performed with the marxs code \citep{2017AJ....154..243G}. However, the setup is chosen to illustrate the concept of a DTRS, not to represent any particular instrument. For example, the position of the diffractive gratings shown leaves valuable mirror area uncovered and is done without considering the size and location of support structures that the gratings can be mounted on. The grating constant $d$ is chosen to be smaller than feasible for a real instrument. Static images of the 3D ray-traces are embedded in the article, but we encourage the reader explore the interactive 3D models embedded in the electronic version of the publication.

The points $\vec p$ on the surface of a torus can be parameterized by two angles $\varphi$ and $\theta$.
\begin{equation}
\vec p(\varphi, \theta) = \vec c + R \; \vec e_R(\varphi) + r \; (\vec e_y \sin \theta + \vec e_R(\varphi) \cos \theta)
\end{equation}
where $\vec c$ points to the center of the torus. The axis of symmetry is given by a unit vector $\vec e_y$. We define a vector $\vec e_R(\varphi) = \vec e_x \cos\varphi + \vec e_z \sin\varphi$. $R \vec e_R$ points from the center of the torus to the center of a circle with radius $r$. Sweeping that circle around the symmetry axis forms the torus. If $r < R$, the torus has an inner hole like a doughnut, and if $r > R$ the inner parts of the torus overlap with each other (``spindle torus'').

Figure~\ref{fig:sketch} shows sketches of the geometry in the plane spanned by the optical axis and the symmetry axis of the torus. Details will be discussed in the following sub-sections.

\begin{figure*}
    \plotone{sketch}
    \caption{\emph{left:} Sketch of the on-axis Rowland geometry in the plane of the optical axis and the symmetry axis of te torus. Photons arrive from the left. The location of the mirror is shown as blue box. The optical axis (solid blue line) passed through the center of the Rowland circle (blue square) and is perpendicular to the symmetry axis of the torus (blue long-dashed line). The two blue circles show the cut through the 3D torus in this plane; they are connected by blue dotted lines to help visualize the 3D position of the torus.
    \emph{center:} Tilted Rowland torus. The optical axis of the blue channel is offset from the center of the Rowland circle, but intersects the symmetry axis of the torus in the focal point $F$. $\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus to the center of the Rowland circle. $\beta$ is the angle from optical axis to the line connecting that intersection to the ``hinge'', the other point where symmetry axis of the torus and the Rowland circle intersect in this plane.
    \emph{right:} Layout with two Rowland tori, tilted in opposite direction. In the plane spanned by both optical axes (which are parallel) and the symmetry axis of the torus, the two tori overlap in exactly one circle, the Rowland circle.
        }
    \label{fig:sketch}
\end{figure*}

\subsection{One Rowland Torus with central optical axis}
\label{sect:onetorus}
For a single Rowland torus with no tilt, the optical axis passes through the center of the Rowland circle and $r=R$ as shown in the left panel of figure~\ref{fig:sketch}. Figure~\ref{fig:3d:single} shows a 3D view of a ray-trace. Since the diffraction gratings are symmetric with respect to the optical axis, positive or negative diffraction orders (up or down in figure~\ref{fig:sketch}) are equally likely, and the diffracted photons will be symmetric with respect to the focal point. To avoid the same wavelength to fall in a chip gap on the detector for both the positive and the negative diffraction orders, it is useful to position the CCDs with an offset, so that the chip gaps are not symmetric.
\begin{figure}
    \plotone{ifigs/single}
    \caption{3D image of the ray-trace for a source with a continuum spectrum over a limited bandpass. Rays are shown from the entrace aperture on (circle on the right). They pass a focussing mirror represented by a green box and get diffracted by gratings (white squares). They are detected by CCDs (yellow). Rays are colored by the diffraction order: Light purple (middle) are zero order photons. A section of the Rowland torus is shown as transparent gray. The big square at the left indicates the position of the focal plane. Rays are traced past the detectors to the focal plane to show how the CCDs on the Rowland circle capture the rays when the dispersion width is smallest - past the CCDs the rays widen in dispersion direction again, but still become narrower in cross-dispersion direction. This figure is interactive in the online version, allowing the reader to pan, zoom, and rotate to see those details and inspect the rays from all angles. (Use the mouse to rotate, ``ctrl'' and mouse to pan, mouse wheel to zoom, and double-click to set the center of rotation. On other devices, other native controls might be available, e.g.\ two-finger touch to zoom on touchscreens.)
        }
    \label{fig:3d:single}
\end{figure}
%\begin{figure}
%  \begin{interactive}{js}{file_wth_arll_components.tar/zip}
%  figure call (e.g.\plotone, \includegraphics, etc.)
%  \end{interactive}
%  \caption{Description of the example figure and the interactive portion which includes how to use the interactive functionality.}
%\end{figure}

@@

%alpha, beta, equations. notation different from Ralf.

\begin{figure*}
    \plotone{ondetector}
    \caption{Distribution of photons on the detectors for different designs. Photons are colored by channel and the position of the zero(s) order(s) is highlighted with red boxes. Note that the dispersion and cross-dispersion directions are not to scale.
        }
    \label{fig:fish}
\end{figure*}

\citet{doi:10.1117/12.856482} suggested tilting the torus with respect to the optical axis when using CAT gratings.
This tilt can be described by two angles $\alpha$ and $\beta$ as shown in the middle panel of figure~\ref{fig:sketch}. $\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (labelled ``A'' in the figure) and the focal point of the mirror (``F''). We call the distance between those two points $a$, which is chosen to be slightly shorter than the focal length $f$ of the mirror to allow for space to mount the gratings behind the mirror.
In this configuration, the torus has $R < r$ there are two points where the torus overlaps itself in the plane spanned by the optical axis and the symmetry axis of the torus (the plane shown in figure~\ref{fig:sketch}). The first of those intersection points is F. We call the second point ``Hinge''.
$\beta$ is the angle from optical axis to the line connecting A and the Hinge. With that we can derive the radius of the Rowland circle to
$$r = \frac{a}{2}\cos \alpha.$$
 If the focal point F is the origin of the coordinate system, then the center of the Rowland circle (blue squares in left blue circles in the panels in figure~\ref{fig:sketch}) is at location
 \begin{equation}
    \begin{pmatrix} x \\ y \\ z \end{pmatrix} =
    \begin{pmatrix} 0 \\ \frac{a}{2} \\  \frac{a}{2}\tan\alpha \end{pmatrix}.
 \end{equation}

Since A, F, and the Hinge are by definition all on the Rowland circle, their distance to this point is $r$. With this symmetry, we can obtain the angle of the axis of the torus with respect to the case where the axis of the torus is perpendicular to the optical axis (left panel of figure~\ref{fig:sketch}). This angle is $\beta-\alpha$. With that, we can derive $R$, the distance between the center of the Rowland circle and the center of the torus, to
\begin{equation}
    R = r \sin\left(\frac{\pi}{2} - \beta\right),
\end{equation}
which places the center of the torus at
\begin{equation}
    \begin{pmatrix} x \\ y \\ z \end{pmatrix} =
    \begin{pmatrix} 0 \\ \frac{a}{2}-R\cos{\beta-\alpha} \\ \frac{a}{2}\tan{\alpha}+R\sin{\beta-\alpha} \end{pmatrix}.
\end{equation}

CAT gratings are blazed and thus deviate more from the Rowland-torus than a grating that is placed perpendicular to the incoming rays. Figure~\ref{fig:3d:tilted} shows this layout in 3D, but with a very large blaze angle for clearity; for real CAT gratings typical blaze angles are 1-2~deg. Tilting the Rowland torus by about twice the blaze angle reduces this effect. Also, CAT gratings disperse predominantly in one direction. However, the Rowland geometry optimizes the spectral resolving power by minimizing the spread of the photons in cross-dispersion direction; detectors are located on the Rowland circle, which represents the position of minimal spread in dispersion direction. This is not the same location as the imaging focal plane, which is defined by minimizing the total size of the PSF. At the focal point, the Rowland circle and the imaging focal plane match. Figure~\ref{fig:fish} shows how the dispersed photons spread out in cross-dispersion direction, roughly resembling the outline of a fish. With a tilted Rowland torus, there is a second intersection point between the spectral and the imaging focus where the cross-dispersion width becomes small (the point between the body and the tail of the fish, about +300 in dispersion direction in the second panel of figure~\ref{fig:fish}). This location can be chosen to be roughly in the center of the dispersed spectrum and thus limit the cross-dispersion profile to reduce the area of the detector that contributes to the extracted signal and thus the background. We note that the width in cross-dispersion direction is much larger for larger diffraction angles and, in real instruments, will be smaller than shown in the figure.

\begin{figure}
    \plotone{ifigs/tilted}
    \caption{Ray-trace like figure~\ref{fig:3d:single}, but for a tilted Rowland torus setup. See figure~\ref{fig:3d:single} for an explanation of the elements shown in the image and the interactivity in the online version.
        }
    \label{fig:3d:tilted}
\end{figure}

A tilted Rowland can also be combined with sub-aperturing to increase the spectral resolving power as discussed in figure~\ref{fig:subaperture}. In this case, not all of the fish shape is filled with photons. This can be seen by comparing the dark purple points in the second, third, and forth panel in figure~\ref{fig:fish}. The Rowland torus for this channel has the same tilt and dimensions in all three designs, but in the second panel, the mirror fills the full circle, in the third panel it has just two wedges, and in the forth panel just one. In either case, the spectral extraction region used in data analysis should take this shape into account to minimize detector background.

\subsection{Double tilted Rowland Torus}
\begin{figure}
    \plotone{ifigs/tilted_double}
    \caption{Ray-trace like figure~\ref{fig:3d:tilted}, but with channels split on two separately tilted Rowland tori. See figure~\ref{fig:3d:single} for an explanation of the elements shown in the image and the interactivity in the online version.
        }
    \label{fig:3d:tilted_double}
\end{figure}
